SPERF a scalability profiler for shared-memory parallel applications

### 2022-04-05 thesis/pascalanalyzer/pascalanalyzer.tex
first version
add journals and restructured

\section{Pascal Analyzer}

Run application collecting information while executing a program with
different configurations of cores, frequencies, inputs arguments.

\subsection{Architecture}
\begin{figure}
	\includegraphics{pascalanalyzer/arch2}
\end{figure}
\subsection{Modules description}
PascalRun: Provie a monitor class to collect information while executing  a program 
with different configurations of cores, frequencies, inputs arguments. The information collected can be via sensor specified by the user,  instrumented regions (source code and openmp binaries) and the entire  program execution time. The data is stored and processed with PascalData  module that can resume all on a DataArray (xArray module) with times, speedups or efficiency.

PascalData: Load saved data from pascalrun and can convert and process to varius formats.

PascalModel: Create AI and Analytical models from pascal data

Instrumenter: Instrument source code creating marks at parallel regions

libinst: Instrument binary code creating marks at parallel regions

cpufreq: resposible of CPU control

IPMI, RAPL, RAPL\_sample, ProcList: Sensors

\subsection{Features}

\subsubsection{Options}
\begin{outline}
	\1 list of cores
	\1 list of frequencies
	\1 list of inputs
	\1 list of sensors
		\2 ipmi
		\2 rapl (sysfs)
		\2 rapl (perf)
		\2 process cores monitor
	\1 number of repetitions
	\1 choice of CPU governor
	\1 idle time between task
	\1 enable/disable hyperthread
	\1 enable/disable cores
	\1 collect times from marks manually inserted by the user in the source code
	\1 collect times from parallel regions using binary instrumentation (openmp)
	\1 list of sensors
	\1 instrument source code
	\1 output file name
	\1 verbosity level
		\2 0 nothing
		\2 1 only display run configuration
		\2 2 display run configuration and sensors data
		\2 3 display everything
\end{outline}

\subsubsection{Execution}

Examples execution using command line
\begin{lstlisting}
pascalanalyzer ./myapp --bin --idle_time 5 --cores 1,4  
		--frequencies 3000000,2800000 --verbosity 3

pascalanalyzer ./myapp -c 1,32 -v 2 -o myoutput.json
\end{lstlisting}

Example execution python
\begin{lstlisting}
from pascalanalyzer.pascalrun import PascalRun
from pascalanalyzer.sensors.rapl import RAPL
rapl_sensor = RAPL()

app = PascalRun(
	application="myapp",
	repetitions=5,
	sensors=[rapl_sensor],
	idle_time=30,
	governor="userspace",
	disable_cores=True,
	disable_hyperthread=True,
)
app.run(
	cores=[1,16,32],
	frequencies=[3000000,2800000],
	inputs=[["arg1,"arg2"], ["arg3,"arg4"]]
)
app.save("output_file")
\end{lstlisting}

Example model creating python
\begin{lstlisting}
from pascalanalyzer import PascalData
from pascalanalyzer.pascalmodel import *

data = PascalData(filename)
df = data.energy()

opt = LeastSquaresOptmizer("model= lambda x, N, p, f:"
                           "x[0]*f**2*p+x[1]/(f*p)*N**x[7]+x[2]*f**2"
                           "+x[3]*p+x[4]/f*N**x[6]+x[5]", 8)

pascalmodel= create_model(df,inputs=["input","cores","frequency"],
                             output=["ipmi_energy"], model= opt,
                             config=data.config, train_sz=0.9)
\end{lstlisting}

\subsection{Data structure}

\subsubsection{Proposals}
\subsubsection{neasted dictonaries}
\begin{lstlisting}
{
	key1:
	{
		key2:
		{
			key3:
			{
				data1: val1,
				data2: val2
			},
			...
		},
		...
	},
	...
}
...
\end{lstlisting}

\subsubsection{index dictonary}
\begin{lstlisting}
{
"key1;key2;key3;...": [
	{
		data1: val1,
		data2: val2
	},
	...
],
}
...
\end{lstlisting}

\subsubsection{Benchmarking}

{\centering

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 10.00 b

\begin{tabular}{llllr}
	\toprule
	{} & Storage &    Neasted &      Index &  Neasted/Index \\
	\midrule
	0 &     RAM &    2.57 Mb &    4.73 Mb &       0.542656 \\
	1 &   JSON  &  616.52 Kb &    1.26 Mb &       0.490468 \\
	2 &     PKL &  380.99 Kb &    1.21 Mb &       0.315070 \\
	3 &      GZ &  221.96 Kb &  270.36 Kb &       0.820989 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 100.00 b

\begin{tabular}{llllr}
	\toprule
	{} & Storage &  Neasted &    Index &  Neasted/Index \\
	\midrule
	0 &     RAM &  4.73 Mb &  6.89 Mb &       0.686078 \\
	1 &   JSON  &  2.78 Mb &  3.42 Mb &       0.812560 \\
	2 &     PKL &  2.54 Mb &  3.37 Mb &       0.754198 \\
	3 &      GZ &  1.82 Mb &  1.87 Mb &       0.968731 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 1.00 Kb

\begin{tabular}{llllr}
	\toprule
	{} & Storage &   Neasted &     Index &  Neasted/Index \\
	\midrule
	0 &     RAM &  26.33 Mb &  28.49 Mb &       0.924100 \\
	1 &   JSON  &  24.38 Mb &  25.02 Mb &       0.974398 \\
	2 &     PKL &  24.22 Mb &  25.04 Mb &       0.966930 \\
	3 &      GZ &  17.58 Mb &  17.64 Mb &       0.996294 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 10.00 Kb

\begin{tabular}{llllr}
	\toprule
	{} & Storage &    Neasted &      Index &  Neasted/Index \\
	\midrule
	0 &     RAM &  242.33 Mb &  244.49 Mb &       0.991156 \\
	1 &   JSON  &  240.38 Mb &  241.02 Mb &       0.997343 \\
	2 &     PKL &  240.24 Mb &  241.07 Mb &       0.996565 \\
	3 &      GZ &  175.11 Mb &  175.07 Mb &       1.000198 \\
	\bottomrule
\end{tabular}

Keys: [32,10,10,10,10] 320.00 Kb  Data bytes: 1.00 b

\begin{tabular}{llllr}
	\toprule
	{} & Storage &    Neasted &     Index &  Neasted/Index \\
	\midrule
	0 &     RAM &   13.38 Mb &  37.27 Mb &       0.358895 \\
	1 &   JSON  &    5.23 Mb &  13.98 Mb &       0.373866 \\
	2 &     PKL &    1.49 Mb &  12.70 Mb &       0.117686 \\
	3 &      GZ &  537.08 Kb &   1.24 Mb &       0.431880 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 10.00 b

\begin{tabular}{llrr}
	\toprule
	{} &      Time &   Neasted &  Index \\
	\midrule
	0 &  Creation &  0.073802 &    0.0 \\
	1 &     Acess &  0.000000 &    0.0 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 100.00 b

\begin{tabular}{llrr}
	\toprule
	{} &      Time &   Neasted &  Index \\
	\midrule
	0 &  Creation &  0.426866 &    0.0 \\
	1 &     Acess &  0.000000 &    0.0 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 1.00 Kb

\begin{tabular}{llrr}
	\toprule
	{} &      Time &   Neasted &  Index \\
	\midrule
	0 &  Creation &  4.162496 &    0.0 \\
	1 &     Acess &  0.000000 &    0.0 \\
	\bottomrule
\end{tabular}

Keys: [16,10,5,3,10] 24.00 Kb  Data bytes: 10.00 Kb

\begin{tabular}{llrr}
	\toprule
	{} &      Time &    Neasted &  Index \\
	\midrule
	0 &  Creation &  42.476255 &    0.0 \\
	1 &     Acess &   0.000000 &    0.0 \\
	\bottomrule
\end{tabular}

Keys: [1]*1000 1.00 Kb  Bytes: 10.00 Kb

\begin{tabular}{llllr}
	\toprule
	{} & Storage &    Neasted &     Index &  Neasted/Index \\
	\midrule
	0 &     RAM &  314.94 Kb &  19.23 Kb &      16.380038 \\
	1 &   JSON  &   23.89 Kb &  18.90 Kb &       1.264328 \\
	2 &     PKL &   23.91 Kb &  18.92 Kb &       1.263904 \\
	3 &      GZ &   10.34 Kb &  10.14 Kb &       1.019525 \\
	\bottomrule
\end{tabular}

Keys: [1]*1000 1.00 Kb  Data bytes: 10.00 Kb

\begin{tabular}{llrr}
	\toprule
	{} &      Time &   Neasted &  Index \\
	\midrule
	0 &  Creation &  0.003023 &    0.0 \\
	1 &     Acess &  0.000000 &    0.0 \\
	\bottomrule
\end{tabular}

}

From the tests, we conclude that index dictionary is a better option for speed and nested dictionary for space, although if the data size is significantly higher than the number of keys it doesn't save a lot of space. Because of that, we choose to work with index.

\subsubsection{Structure}
The data structure holds

\begin{lstlisting}
{
    "config": {
        "pkg": "./openmp_simple_bin",
        "execdate": "29-03-2020_13:07:07",
        "command": "run_test.py TestMonitor.test_valid_combinations",
        "hostname": "MSI",
        "kernel": "Linux-4.4.0-18362-Microsoft-x86_64-with-Ubuntu-18.04-bionic",
        "disable_hyperthread": true,
        "disable_cores": true,
        "idle_time": 0,
        "data_descriptor": {
            "values": [ ... ],
            "extras": {
                "regions": {
                "values": [ ... ]
            }
         },
        "keys": [ ... ]
    },
    "arguments": [ ... ],
    "governor": "userspace"
    },
    "data": {
        "4;1.2e-06;1;1": {
            "regions": {
                "0x1": [
                        [ ... ],
                        ...
                    ],
                ...
            },
            ...,
            "total_time": 0.3075871467590332
        },
        ...
    }
}
\end{lstlisting}

\subsection{Overhead}
\subsubsection{BlackSchoels test}

$number\_of\_tests=96$\\
$max\_number\_threads=32$\\
$type\_of\_step=power$\\
$value\_of\_step=2$\\
$list\_of\_args=\{in\_312K.txt \quad outputFile.txt\}$

\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\textwidth]{pascalanalyzer/figures/black100b}
	\caption{With SPERF}
	\label{fig:black100b}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\textwidth]{pascalanalyzer/figures/black2100b}
	\caption{With Linux Time}
	\label{fig:black2100b}
	\end{subfigure}
\end{figure}

\subsubsection{Freqmine test}

$number\_of\_tests=96$\\
$max\_number\_threads=32$\\
$type\_of\_step=power$\\
$value\_of\_step=2$\\
$list\_of\_args=\{kosarak\_990k.dat \quad 1200 \quad output.txt\}$

\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/freq100b}
		\caption{With SPERF}
		\label{fig:freq100b}
	\end{subfigure}
	%
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{pascalanalyzer/figures/freq2100b}
		\caption{With Linux Time}
		\label{fig:freq2100b}
	\end{subfigure}
\end{figure}

\subsubsection{Intrusiveness measurement}

The statistical result obtained from the test above from Blackscholes and Freqmine was shown in the tables \ref{tab:blackstatistics} and \ref{tab:freqstatistics}.Where we can compare the result mean and standard deviation from two groups one with SPERF and the other without.

\begin{table}[H]
	\centering
	\caption{Statistics for BlackSchoels}
	\label{tab:blackstatistics}
	\begin{tabular}{llll}
		& With Sperf & Without Sperf & Total    \\
		N               & 96         & 96            & 192      \\
		$\sum{X}$       & 100,19     & 103,0134      & 203,2034 \\
		Mean            & 1,0436     & 1,0731        & 1,0584   \\
		$\sum{X^2}$     & 105,5093   & 112,4857      & 217995   \\
		Std.Dev.        & 0,0998     & 0,1431        & 124     
	\end{tabular}
\end{table}

\begin{table}[H]
	\centering
	\caption{Statistics for Freqmine}
	\label{tab:freqstatistics}
	\begin{tabular}{llll}
		& With Sperf & Without Sperf & Total    \\
		N           & 96         & 96            & 192      \\
		$\sum{X}$   & 202,8016   & 205,44        & 408,2416 \\
		Mean        & 2,1125     & 2,14          & 2,1263   \\
		$\sum{X^2}$ & 429,9198   & 440,428       & 870,3478 \\
		Std.Dev.    & 0,1256     & 0,091         & 0,1102  
	\end{tabular}
\end{table}

Given the summary statistics, the calculations of the hypothesis test are shown in tabular form in the tables \ref{tab:blackAnova} and \ref{tab:freqsAnova}.

\begin{table}[H]
	\centering
	\caption{Anova f test}
	\label{tab:blackAnova}
	\begin{tabular}{lllll}
		Source             & SS     & df  & MS     &             \\
		Between-treatments & 0,0415 & 1   & 0,0415 & F = 2,72671 \\
		Within-treatments  & 2893   & 190 & 0,0152 &             \\
		Total              & 2,9345 & 191 &        &            
	\end{tabular}
\end{table}

\begin{table}[H]
	\centering
	\caption{Anova f test}
	\label{tab:freqsAnova}
	\begin{tabular}{lllll}
		Source             & SS     & df  & MS     &             \\
		Between-treatments & 0,0363 & 1   & 0,0363 & F = 3.01548 \\
		Within-treatments  & 2,2844 & 190 & 0,012  &             \\
		Total              & 2,3206 & 191 &        &            
	\end{tabular}
\end{table}

If we take a significance level of 5\% the critical value of F with 1 and 190 degrees of freedom on numerator and denominator respectively is 3.8909. Size booth values of F calculated is lower then the critical there is no significant statistical difference on the mean value according to ANOVA.

%\begin{figure}[H]
%	\includegraphics[width=7cm,height=4.5cm]{ftest}
%	\caption{F distribution}
%	\label{fig:Fdist}
%\end{figure}

\subsubsection{Sampling}

Energy estimation method for power samples
Time difference
\begin{equation}
	E= \sum_i^{N-1}(t_{i+1}-t_i)p_i+s_N(t_{tot}-t_N)
\end{equation}

Constant sample time
\begin{equation}
E= s_r\sum_i^{N}p_i=TP_{avg}
\end{equation}

Error due to sampling
\begin{itemize}
	\item Time difference $\approx$ 4\%
	\item Constant sample time 	$\approx$ 1\%
\end{itemize}

\subsection{Database}

\begin{outline}
	\1 Advantages
		\2 No need to organize the files
		\2 Faster and easier queries
		\2 Shared data across programs and research
	\1 Disadvantages
		\2 Need of db server
		\2 Need of network
		\2 Non local server maybe slow ( need to cache )
\end{outline}

>>>>

\section{Pascal Analyzer} \label{sec:pascal_analyzer}
High-performance computing systems have become increasingly dynamic, complex, and unpredictable. To help build software that uses full-system capabilities, performance measurement and analysis tools exploit extensive execution analysis focusing on single-run results. Despite being effective in identifying performance hotspots and bottlenecks, these tools are not sufficiently suitable to evaluate the overall scalability trends of parallel applications. Either they lack the support for combining data from multiple runs or collect excessive data, causing unnecessary overhead. 
In this work, we present a tool for automatically measuring and comparing several executions of a parallel application according to various scenarios characterized by the input arrangements, the number of threads, number of cores, and frequencies.
%In this work, we present a tool for measuring and comparing several executions of a parallel application automatically with different input arrangements, number of threads, cores, and frequencies to understand its scalability by observing how it uses the available computational resources across configurations. 
Unlike other existing performance analysis tools, the proposed work covers some gaps in specialized features necessary to better understand computational resources scalability trends across configurations. 
%The proposed work covers some gaps and unspecialized features of existing performance analysis tools. 
%The tool features automatic instrumentation and direct mapping of parallel regions, accuracy-preserving data reductions, and ease of use to compare multiple configurations for improving analysis and productivity. 
In order to improve scalability analysis and productivity over the vast spectrum of possible configurations, the proposed tool features automatic instrumentation, direct mapping of parallel regions, accuracy-preserving data reductions, and ease of use.
%In addition, it aims at parallel scalability, instead focus on single-run performance, aggregating a minimal level of intrusion (<1\% on average), which is fundamental for accurately understanding the behavior and scalability trends of the parallel application.
As it aims at accurately understanding scalability trends of parallel applications, detailed single-run performance analyses show minimal intrusion (less than 1\% overhead).

\subsection{Introduction} \label{sec:introduction_pascal}

Developing parallel programs capable of exploiting the computational power of high-performance systems is as well-known as it is challenging~\cite{Huck2007, Islam2019, Weber2019}. 
During the development process, the code optimization step is a fundamental part of the software construction strategy and is supported by performance measurement and analysis tools~\cite{Bergel2019, Weber2019, Huck2005, Geimer2010, Shende2006, Adhianto2010, Miller1995, Galobardes2015, Pillet2007, Islam2019}. The main objective of these tools is to help developers understand the execution characteristics, allowing the identification of bottlenecks and behavioral phenomena that compromise the program’s efficiency, thus guiding possible improvements~\cite{Brink2020, Huck2007}.

Nowadays, due to the complexity of parallel systems, correctly identifying and locating performance and scalability bottlenecks depends on the developer's ability to compare several measurements in different execution configurations~\cite{Bergel2019, Silva2018}. This investigative process tends to be tedious and complex, requiring in-depth knowledge of the problem domain, parallel systems, and measurement and analysis tools. 
%Mastering the analysis tools becomes necessary mainly because they differ from each other in terms of measurement strategy, the collected metrics, and the many points and focus of observation. 
Mastering the analysis tools is necessary because they differ from each other in terms of measurement strategy, metrics collected, and the many points and focus of observation.
Even though some solutions propose organizing and combining metrics from different sources, these distinct characteristics can compel developers to exploit more than one tool depending on their analysis~interests.

%Despite their differences, every existing tool has its own contributions in helping to understand the inner working of a program, and some can present a vast and varied set of metrics with a high degree of detail. 
Every existing tool has its own contributions in helping to understand the inner working of a program. Some can present a vast and varied set of metrics with a high degree of detail. 
However, developers may have difficulty using these tools when their interest is in analyzing parallel scalability. 
%The main challenge arises from the focus on investigating a single run, because scalability analysis requires contrasting runs of different configurations. 
The main challenge arises from the emphasis on studying a single run because scalability analysis requires contrasting runs of different configurations.
%In this sense, it tends to be more significant to measure and compare only a few data across executions of multiple configurations than to collect a vast amount of finer-grain data of single-configuration runs.
In terms of productivity, measuring and comparing only a few data between runs of multiple configurations tends to be more significant to scalability analysis than collecting a vast amount of finer-grain data of single-configuration runs.
Furthermore, due to the number of measurements, there is a natural overhead associated with the execution of the analysis tool itself. Some works indicate this overhead can reach 40\% of the program runtime, which influences and damages the observation of its efficiency variation~\cite{Eriksson2016}.
As such, a measurement and analysis approach that uses only the data needed to understand the application's efficiency variation can be advantageous. First, because measuring and collecting only the necessary data limits the degree of intrusion and, second, because this strategy makes the analysis tool simpler and easier to use. 
%Finally, the tool's availability can also be a drawback as some apply to specific compilers and execution frameworks. Indeed, the framework version is a limitation for some traditional tools like HPCToolkit~\cite{Adhianto2010} and TAU (Tuning and Analysis Utilities)
%MDPI: Is this an abbreviation? If affirmative, please add the explanation. Please carefully check and confirm -- INCLUDED ON THE NEXT PARAGRAPH
~\cite{Shende2006}.

From this context, we present a measurement tool that focuses on analyzing the parallel scalability of programs. It primarily provides runtime and power consumption measures in an approach that favors inspection of the overall behavior of the program before starting a more in-depth investigation of points that naturally require more detail, time and effort.
The work explores and combines strategies already used in other solutions, such as tracing, post-mortem analysis, and automatic instrumentation based on shared libraries. Additionally, it includes an instrumentation model that automatically relates parallel code regions to analysis regions, using a measurement strategy based on the intrinsic and primary behavior of the parallel frameworks. Due to this approach, the tool allows analyzing programs developed for shared-memory environments regardless of the compiler or framework versions. 
The framework version, which in some cases also requires specific compiler versions, is a limitation for some traditional tools like HPCToolkit~\cite{Adhianto2010} and TAU (Tuning and Analysis Utilities)~\cite{Shende2006}.

The proposed tool uses a tracing-based measurement technique to identify and measure the boundaries of the analysis regions. The tracing approach supports a hierarchical region analysis model allowing users to see how the efficiency of inner parts can impact the program's overall scalability. In addition, the tool uses an aggregation feature to reduce the volume of data produced while preserving accurate analysis capability with minimal overhead. The tool also includes usability features that reduce the developer's efforts during the measurement and analysis process. These features allow one to automate the process of running and merging data collected from multiple configurations, bypassing manual efforts and avoiding the drawbacks of non-computer-oriented analysis. 
%The tool does not display visuals elements natively. Still, its interface allows the collected data to be exported and viewed from the control terminal in tabular form. Data exported via the interface are organized into data frames that can be interpreted by other tools or even rendered directly from graphic libraries.
The collected data, available from the terminal in tabular form, can be exported as data blocks that can be interpreted by other tools or even rendered by graphics libraries to facilitate the visualization and data analysis.

%The proposed tool is an alternative for realizing parallel scalability analysis more efficiently than single-run-centric performance measurement and analysis tools. 
We offer an alternative tool for realizing parallel scalability analysis more efficiently than single-run-centric performance measurement and analysis tools. In addition to that, our framework can also be used to promote energy-aware software; methods that rely on accurate and efficient energy collection could benefit from energy consumption data for the entire program and specific regions of the application \cite{10.1007/978-3-319-58667-0_22,10.1007/978-3-319-58667-0_21,7016382,10.1145/3321551}. To summarize, the main contributions of this work are:

\begin{itemize}
	\item A lightweight automated process of measuring and comparing runs with different configurations.
	\item Transparent linking of parallel code sections to analysis regions.
	\item Hierarchical views and degrees of efficiency for the inner parts of a program.
	\item The imposed overhead is low and minimally intrusive, mainly because of the measurement approach and the number of metrics collected.
	\item It provides energy consumption measures for the different configuration parameters. 
\end{itemize}

The following sections describe how the tool is built and used for performance analysis. Section \ref{sec:related_work_pascal} presents the related works. Section \ref{sec:design_and_features_pascal} describes the tool architecture and goals; it explains the main features, usage, and collected data. The experimental results are presented in Section \ref{sec:validation_the_framework}. Finally, the contributions are summarized in Section \ref{sec:conclusions_pascal}, with an outlook of future works.

\subsection{Related Works} \label{sec:related_work_pascal}

Performance analysis aimed at code optimization is an essential activity for developing parallel applications with scalable performance. For this purpose, performance analysis tools, like Caliper~\cite{Boehme2016}, HPCToolkit~\cite{Adhianto2010}, Scalasca~\cite{Geimer2010}, TAU~\cite{Shende2006}, Vampir~\cite{Weber2019}, and VTune~\cite{VTune}, are useful because they help developers identify where the application uses the available computing resources inefficiently through collecting detailed data of its execution. Such tools differ from each other mainly in terms of data measurement and analysis strategy: tracing or profiling, post-mortem or real-time; individual or comprehensive observation focus; and for providing or not providing visual elements that facilitate the judgment of collected data. Other solutions, like the DASHING framework~\cite{Islam2019} and the HATCHET library~\cite{Brink2020}, allow inspection of performance data from multiple sources. In this case, the goal of DASHING and HATCHET is to provide a more robust data set to support the analysis process.

In the performance analysis process, observing the parallel program's execution as a sequence of events representing significant activities is essential to understand its behavior~\cite{Pantazopoulos1997}. Events are basic units of the analysis process, and the way they are observed influences strategies for collecting performance data. Profiling-based analysis tools collect information from events that occurs in the program execution and commonly operate by statistical sampling using interruptions. These interruptions can be caused, for example, by periodic breaks or hardware events. Interruptions allow checking the system state and the information collected depends on the focus of observation. Profiling-based tools use statistical techniques to describe the program behavior in terms of aggregate performance metrics. They usually ignore the chronology of events, but are known to be useful for identifying, for example, load imbalance, high communication time, or excessive routine calls. Paradyn~\cite{Miller1995}, and Periscope~\cite{Gerndt2010} are tools that adopt this strategy.

On the other hand, tracing-based analysis tools collect performance data from events that occur when the program takes over a particular state. Tracing can provide valuable information about the time dimension of a program, allowing users to check when and where transitions of routines, communications, and specific events occur. This measurement strategy tends to be more invasive and intrusive, generating a larger dataset than the profiling-based approach. Vampir~\cite{Weber2019} and Paraver~\cite{Labarta2005} are examples of this group. Some tools like HPCToolkit~\cite{Adhianto2010}, Scalasca~\cite{Geimer2010}, TAU~\cite{Shende2006} and VTune~\cite{VTune} support both profiling and~tracing.

%In this work we propose tracing to measure the program and identify when parts of its code were executed. 
In this work we propose tracing to identify when parts of a code were executed. 
%To mitigate some of the disadvantages usually related to tracing, the tool implements features such as automatic instrumentation mode and aggregation resource, in addition to adopting an analysis based only on time. 
To mitigate some of the disadvantages related to tracing, in addition to adopting a time-based analysis, the tool implements features such as automatic instrumentation mode and resources aggregation.
%The automatic instrumentation mode makes it possible for users to analyze the program's execution without modifying the compiled code. 
Our automatic instrumentation mode allows users to analyze program execution without modifying the source or the executable.
Although many other analysis tools use the automatic instrumentation mode, this work presents a different way of mapping and measuring the code parts. 
%In our proposal, parallel regions are the main focus of analysis, and there is no limitation on the version of OpenMP used in the development, as in tools such as HPCToolkit~\cite{Adhianto2010} and TAU~\cite{Shende2006}. 
In our proposal, parallel regions are the main focus of analysis. 
%The strategy is efficient in identifying the scalability trend of parts or the whole parallel program in execution.
The strategy is effective enough to identify the scaling trends of parts or of the entire running parallel program.
%Furthermore, collecting only data related to runtime reduces the associated overhead with the measurement process. 
Furthermore, collecting only specific runtime data reduces the overhead associated with the measurement process.
As with other similar tools, such as HPCToolkit~\cite{Adhianto2010} and TAU~\cite{Shende2006}, there is no limitation on the version of OpenMP~used.

Runtime overhead is an attribute associated with the set of additional instructions that are executed to collect the program measurements. The time to perform this ``extra code'' varies in line with the tool and is directly related, in quantity and degree of detail, to the aspects it measures. 
Limiting this overhead is crucial because it can compromise the understanding of program behavior, and divert optimization efforts to less effective points. 
According to some comparative and practical studies in this regard, tests on traditional tools have shown a runtime overhead ranging from 2\% to 40\%~\cite{Eriksson2016}.
%Fundamental to the proposal of this work, the overhead presented by the proposed tool differs according to the program under analysis but can be considered negligible (<1\%) for analyzing scalability trends of parallel applications.
Although different depending on the program analyzed, the overhead resulting from our instrumentation strategy can be considered negligible (less than 1\%) and optimal for analyzing trends in parallel applications.

Several performance analysis tools employ a post-mortem approach. 
%In this kind of approach, the tools collect performance metrics during the program execution, but the data interpretation is carried out only after its execution. 
In this case, the tool performs performance measurements and data collection while the program is running, and then the collected data will be interpreted.
HPCToolkit~\cite{Adhianto2010}, Scalasca~\cite{Geimer2010}, TAU~\cite{Shende2006} and VTune~\cite{VTune} are examples of post-mortem performance analysis tools. 
%Post-mortem performance analysis tools may require the storage of large volumes of performance data, but they are more suitable to provide an overall view of the execution. 
Depending on the type of analysis, these tools may require storing large amounts of data, but they are best suited to provide an overall view of the execution.
%On the other hand, real-time analysis tools enable performance analysis during the execution of the program. 
In contrast, run-time analysis tools perform both operations at run time.
This approach allows detecting waiting states and communication inefficiencies accurately. 
%However, an online analysis environment usually requires the coordinated usage of extra tools, which increases the analysis infrastructure complexity. 
Periscope~\cite{Gerndt2010} and Paradyn~\cite{Miller1995} are examples of tools in this group.
However, run-time analysis generally requires the coordinated action of tools, which increases the structural complexity.
%The need for synchronous initialization, communication between all resources used in the analysis, and the additional load on the network are disadvantages of the real-time approach. 
The need for synchronous initialization and communication between analysis resources as well as their impact on the runtime environment are the drawbacks of this approach.
%In addition, the analysis that depends on data collected at the end of the program's execution, as in the case of scalability analysis, are potentially damaged by the infrastructure overhead and do not, in practice, benefit from characteristics of real-time tools. 
In addition, the analysis, particularly that of scalability, which relies on the data collected at the end of the execution, is potentially degraded by the infrastructure overhead and therefore does not benefit from the characteristics of real-time tools.

\textls[-15]{Energy management solutions can also provide similar features for example GEOPM~\cite{10.1007/978-3-319-58667-0_21},} allows the measurement of time and energy of specific regions. The main advantage that we could not find in any other framework is that our tool can provide specific data of energy consumption of parallel regions in a completely automated way. For example, in GEOPM, it is necessary to instrument the source code to obtain the same result.

%To investigate the scalability trend, it is more important to identify the program's general behavior than its state at a given moment of execution. 
%Although the current performance analysis tools are helpful and allow to identify numerous aspects regarding the program's behavior, none of them presents characteristics directly focused on observing the scalability trend of parallel code. An indispensable feature for this purpose includes automating the program's measurement process in different configurations because a more accurate understanding of scalability requires measuring and comparing multiple code runs.
%Focused on more practical and objective analysis, we propose an alternative tool to observing a parallel program's scalability trend. It collects only information that allows deriving the overall program runtime, or parts of it, according to the user's interest, and includes relevant functionalities to an efficient measurement and analysis process. Among the supported features it provides: the automation of program runs based on different parameters; the generation of data sets that contain metrics with different execution parameters; and the reduction of the generated data volume on disk.

%Although the current performance analysis tools are helpful and allow to identify numerous aspects regarding the program status, to investigate scalability trends, it is more important to identify the behavior of parallel code rather than its state at a given moment of the execution, therefore the post-mortem approach is more suitable. 
By focusing on objective analysis, we offer an alternative tool to observe the scalability trend of a parallel program. Thus, we collect just the information allowing us to infer the overall behavior of the program, or, according to the interest of the user, specifically chosen parts of it, including features relevant to an effective measurement and analysis process. 
To this end, it is essential to also automate the processes to obtain comparative measurements of several runs. 
Supported features are the automation of program executions according to various configuration parameters, including a user-specified granularity, the generation of datasets containing comparative measurements of runtime parameters, and efficient optimization of the amount of produced data.
%In this work, the tool mitigates the disadvantage related to the volume of data set by aggregating the collected data.
Regarding the amount of data collected, our approach uses data aggregation, which facilitates storage and further analysis.

\subsection{Design and Features} \label{sec:design_and_features_pascal}

This section presents the architecture, components, and interconnections used in the design of the proposed tool. It also includes a discussion about main features, instrumentation modes, and output file structure that contains the data measured and collected by the proposed tool.

\subsubsection{Architecture} \label{subsec:architecture_pascal}
The tool we propose to measure, combine and compare multiple runs of a parallel application implements two main concepts to achieve this goal: actuators and sensors.

Actuators characterize the parameters we intend to observe. They are variables representing elements external to the program and which, when changed, can influence aspects such as the performance and efficiency of the running application.~Therefore, analyzing the result of actuators' variation is essential to understand how these elements impact program behavior, especially scalability and power consumption. By default, the tool implements actuators controlling the number of active cores and threads, the program input, and the CPU operating frequency.

The sensor is the concept we use to represent the elements addressed to measure and monitor the variation of actuators. Both sensors and actuators are plugged into the tool core as modules. Currently, the tool implements three types of sensors:

\begin{itemize}
	\item Begin/end: collects data at launch and end of each program run.
	\item Time sample: periodically collects data.
	\item Events-based: collects data when a specific event happens.
\end{itemize}

Currently, there are sensors to measure time, energy, and performance counters. The default sensor is a \textit{begin/end} type used to collect the execution time of the whole application. To measure energy consumption, we developed sampling sensors capable of retrieving data from RAPL (Running Average Power Limit) and IPMI (Intelligent Platform Management Interface), which are interfaces that provide power information from the CPU
%MDPI: Please add the explabnation % added in abbreviations
and the entire system. Apart from that, there are \textit{time sample} and \textit{begin/end} sensors to gather performance counters data.

Measurements that support scalability analysis are taken from event-based sensors. For this, the tool includes markers that trigger events to automatically identify the boundaries of parallel regions defined by developers in codes that use POSIX Threads or OpenMP.
%MDPI: Are these abbreviations? If affirmative, please add the explanations. Please carefully check and confirm % added in abbreviations
This feature is essential for relating parallel code sections to analysis regions and measuring execution times directly from binary code. The triggers for this measurement mode are implemented in a wrapper library and therefore already available in the system. The tool also provides manual marks that can be inserted into the source code to monitor specific parts of the program. Furthermore, any system event can be set as a trigger.
\cref{fig:pascal_architecture} describes the integration of each part of the proposed software. The idea is that actuators and sensors are modular parts that can easily be added or removed. The tool's core is responsible for all operations, launching the application, and data gathering.

\begin{figure}[H]	
	%\includegraphics[width=15 cm]{Definitions/logo-mdpi}
	\includegraphics[scale=0.7]{pascalanalyzer/figures/designfeatures/pascal_philosophy.pdf}
	\caption{Architecture showing the interconnections of the central parts of the tool. Either binary (using the wrapper library) or an instrumented source code, the target application can be launched on the target platform by the tool core following the configuration parameters chosen by the user while deploying actuators/sensors. The resulting data collection is stored in Json file format for post analysis and visualization (GUI). \label{fig:pascal_architecture}}
\end{figure}

\subsubsection{Instrumentation and Intrusiveness} \label{subsec:intrusiveness_pascal}

The instrumentation module is one of the most crucial. In addition to automating the instrumentation, it determines the overload and level of intrusiveness of the instrumentation process. This module is designed to execute the fewest instructions possible in the most optimized manner. Currently, there is instrumentation support for C/C++ languages through shared libraries that work with either automatic or manual instrumentation. Manual instrumentation is preferred when it is necessary to examine the application program in specific code sections, regardless of the type of content.

The manual instrumentation mode is implemented in three routines: one for initialization, another to mark the beginning of the region of interest ({\tt analyzer\_start}), and finally, a routine to mark the end of that region ({\tt analyzer\_stop}). The initialization routine is called when loading the library to create the necessary data structures and set up the data exchange communication. The routines {\tt analyzer\_start} and {\tt analyzer\_stop} identify threads in a region and store timestamps. These routines are implemented in such a way that only one thread at a time writes in a designated position of a two-dimensional array, thus ensuring thread safety and eliminating the necessity of locks.
